---
title: "animated_us_covid_cases_viz"
author: "T Tesfaye"
date: "9/3/2020"
output: 
  html_document:
    toc: true
    theme: journal
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(ggthemes)
library(transformr)

# Parameters
path = "../../tsiontesfaye.github.io/COVID-19/data/nyt-us-states.csv"

#-------------------------------------------------------------------------------
# Code

# Read in the data
covid_states <- read_csv(path)
```

# Intro

I was interested in viewing infection rates in different states since the beginning of the pandemic. This vizualization uses the open-sourced New York Times [covid-19-data](https://github.com/nytimes/covid-19-data) and demonstrates cases in states.


# Visualize Covid Cases per State

```{r}
# Filter states with more than 100,000 cases to be labeled
state_labels <- 
  covid_states %>%
  filter(cases > 100000)
```

Plot the graph anmiated.

```{r}
covid_states %>% 
  ggplot(aes(date, cases, color = state)) +
  geom_line(show.legend = FALSE) +
  #geom_point(show.legend = FALSE) +
  #scale_color_viridis_d() +
  scale_y_continuous(
    label = scales::label_number(scale = 1e-3)
  ) +
  ggrepel::geom_label_repel(data = state_labels, aes(label = state), show.legend = FALSE) +
  #gganimate::transition_time(date) +
  gganimate::transition_reveal(date) + #transition reveal calculates the intermediary values so the line is filled
  gganimate::shadow_mark() + #shadow_mark says leave the previous states on the plot
  theme(legend.position = 'none') +
  theme_minimal() +
  labs(
    title = 'US Covid-19 Cases in: {frame_along}',
    subtitle = "New York became the hotspot starting mid March.\ntate names shown for > 100,000 cases",
    #subtitle = 'Frame {frame} of {nframe}',
    y = "Number of Cases Per State \n (in thousands)",
    caption = "Source: New York Times",
    x = NULL
  )
```

# Conclusion

New York state emerges as the hotspot early on but is recently overtaken by California and Texas. 

# Recommended next steps

Potential next steps include cross-referencing this cases plot with a similar plot for deaths. Standardizing the cases per state population might also give a clearer picture of infection concentration.
