---
title: "Global Covid Data Prep"
author: "T Tesfaye"
date: "`r Sys.Date()`"
output: 
  html_document:
    highlight: haddock
    theme: journal
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
# Libraries
library(tidyverse)
library(dcl)
library(lubridate)
library(rvest)
library(ggrepel)



url_data <- "https://www.worldometers.info/coronavirus/#countries"
css_selector <- "#main_table_countries_today"

file_out <- "../Tableau/global_covid_numbers.csv"
file_out2 <-  "../Tableau/global_covid_nums_modified.csv"
file_out_mortality <- "../Tableau/global_mortality_rate.csv"
file_in <- "../Tableau/global_covid_numbers.csv"
file_out_active_cases_num <- "../Tableau/global_active_cases.xls"
```

## Introduction
We want to understand find out countries most affected by covid-19. We will start by scraping the www.worldometers.info which updates its global covid cases count every day. This data was scraped on July 26, 2020. 

## Scrape the Website

```{r}
#read in the table from the worldometers website

global_covid_numbers <- 
  url_data %>% 
  read_html() %>% 
  html_node(css = css_selector) %>% 
  html_table() %>% 
  as_tibble()
```

Then, write out the table into a csv so that we can parse it back properly converting the characters to numbers


```{r}
global_covid_numbers %>% 
  write_csv(path = file_out)
```

## Wrangle and Output Overall Data
Read in the csv converting the relevant cols to numbers and eliminating the three trailing cols we don't need

```{r}
global_covid_nums_csv <- 
  file_in %>% 
  read_csv(
    col_types = cols_only(
      `Country,Other` = col_character(),
      TotalCases = col_number(),
      NewCases = col_number(),
      TotalDeaths = col_number(),
      NewDeaths = col_double(),
      TotalRecovered = col_number(),
      NewRecovered = col_number(),
      ActiveCases = col_number(),
      `Serious,Critical` = col_number(),
      `Tot Cases/1M pop` = col_number(),
      `Deaths/1M pop` = col_number(),
      TotalTests = col_number(),
      `Tests/1M pop` = col_number(),
      Population = col_number(),
      Continent = col_character()
    ) 
  )
```

Note, this intorduces new parsing errors. However `problems(global_covid_nums_csv)` shows that these errors occured due to an existing N/A in the data so we aren't worried. Hence, rename the columns as desired and output the cleaned up data.

```{r}
global_covid_nums_modified <- 
  global_covid_nums_csv %>% 
  pivot_longer(
    cols = c(TotalCases, ActiveCases, TotalDeaths),
    names_to = c("total_active_deaths"),
    values_to = 'total_active_deaths_num'
  )
  
 
global_covid_nums_modified %>% 
  write_csv(file_out2)
```


```{r}
#overwrite the existing global_covid_numbers csv with this correctly parsed data.

global_covid_nums_csv %>% 
  write_csv(path = file_out)
```

```{r}
# rows to remove
continents = c("North America", "South America", "Asia", "Europe", "Africa", "Oceania", "World")
```



## Extract and Output Mortality Rate

Finding the mortality rate



```{r}
`%notin%` <- Negate(`%in%`)

mortality_rate <- 
  global_covid_nums_csv %>% 
  mutate(mortality_rate = TotalDeaths / TotalCases * 100)


mortality_rate %>% 
  write_csv(file_out_mortality)
```


```{r}
"North America" %notin% continents

global_covid_nums_csv %>% 
  filter(!("Country,Other" %in% continents)) 
```


```{r}
global_covid_nums_csv %>% 
  write_excel_csv(file_out_active_cases_num)
```

## Visualize and Analyze in Tableau

This step will take place in Tableau. 
