---
title: "opportunity_atlas_nyt_viz"
author: "T Tesfaye"
date: "8/26/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Libraries
library(here)
library(tidyverse)
library(tidycensus)
library(sf)
library(stringr)



# Parameters
input_san_jose_atlas_data <- here::here("data/san_joseatlas_tidy_challenge_Jenny.rds")
output_san_jose_atlas_census_data <- here::here("data/san_jose_atlas_census_data.geojson")

input_entire_us_oppAtlas <- here::here("data/entire_us_tract_outcomes_oppAtlas.csv")
output_entire_us_oppAtlas <- here::here("data/entire_us_atlas_census.geojson")


#-----------------------------------
# Code
san_jose_atlas_data <- read_rds(input_san_jose_atlas_data)
entire_us_atlas <- read_csv(input_entire_us_oppAtlas)
```



## Project Description
_Project Inspired by Jenny Nova_

### Part I: Get Census Data

Tidycensus is a package that provides R users with access to the US national census. It has two major functions:

1. `get_decennial()` which is used to access the 1990, 2000, and 2010 decennial US Census APIs, and 
2. `get_acs()`, which is used to access the 5-year American Community Survey APIs

For this project, we will focus on the `get_acs()` function. This function requires a few key arguments to return the desired dataset. The two that need explanation are:

* `geography`: the specification of the US geographical designation for the desired data. See [here](https://walker-data.com/tidycensus/articles/basic-usage.html) to find the various options.
* `variables`: There are over 22,000 variable IDs across different census files. We want to analyze the median household income in `2015`. Hence, we will load the variables for `2015` from the `acs5` file and search the resulting tibble for "Median Household Income." This results in the code: `B19013_001`. See [here](https://walker-data.com/tidycensus/articles/basic-usage.html) for more detail.

```{r}
vars <- load_variables(2015, "acs5", cache = TRUE)
```

Now we are ready to load the median income data for Santa Clara in 2015.

```{r}
tract_boundaries <- 
  tidycensus::get_acs(
  geography = "tract",
  variables = "B19013_001",
  year = 2015,
  state = "CA",
  county = "Santa Clara",
  geometry = TRUE
)
```

### Part II: Merge Census with Opportunity Atlas Data

Because we want to visually represent the income information on a map, and the income information is on the atlas dataset while the geographic information is on the census dataset, we want to join these two datasets.

```{r}
income_with_boundaries <- 
  tract_boundaries %>% 
  left_join(atlas_data, by = "GEOID")
```

### Part III: Output the Data for Mapbox
Mapbox, the interactive visualizing platform, requires its input to be in a `geojson` format. Hence, we will write the above dataframe into a geojson.

```{r}
income_with_boundaries %>% 
  write_sf(output_san_jose_atlas_census_data)
```


### Part IV: Determine the Colors for Mapbox


We want to use the Red, Yellow, Blue color tone as used by the NYT article.
First find all the possible colors

```{r}
RColorBrewer::display.brewer.all()
```

```{r}
RColorBrewer::brewer.pal(n = 11, "RdYlBu")
```

We'll now break the income data based on this breakdown based on percentages from 0 to 100% at every checkpoint at 10%s.

```{r}
quantile(atlas_data$kid_indiv_income, probs = seq(0, 1, 0.1))
```


We'll use these values to color Mapbox.


# Mapbox output

Here is a link to the interactive mapbox output. For ease of accessibility, we will also save a static version of the map here.

```{r}
# Parammeters

zoom <- 9.3
width <- 1200
height <- 800
latitude <- 37.2138
longitude <- -121.7106
access_token <- 
'pk.eyJ1IjoidHRlc2ZheWUiLCJhIjoiY2s3Zm12ZXFwMDRrcDNsb2RrNDBoMXk3aSJ9.Lrsf4_yVFFwblZNM_eGJuw'
style_id <- 'ckec0p24w1p1u1an16l2ngbpt'
username <- 'ttesfaye'
  

mapbox_api_call <- 
  api_call <- 
  str_glue(
    "https://api.mapbox.com/styles/v1/{username}/{style_id}/static/",
    "{longitude},{latitude},{zoom}/{width}x{height}?access_token={access_token}"
  )

mapbox_api_call
```

Save and display this image as png.

```{r}
mapbox_img_output <- here("/data/san_jose_oppAtlas.png")
download.file(url = mapbox_api_call, destfile = mapbox_img_output)
```

