---
title: "reproducing_fed_rate_plot"
author: "T Tesfaye"
date: "9/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Library
library(tidyverse)
library(here)
library(readxl)
library(janitor)
library(lubridate)



# Parameters
input_fed_rate <- here::here("data/fed_funds_rate_1985-2020.xls")


#---------------------------------------------------------
# Code

# Read in the sheet titled FRED Graph and skip the first 10 lines which are notes
fed_rate <- 
  read_xls(input_fed_rate, skip = 10, sheet = 'FRED Graph')
```
## Intro

In August 2020, the economic writer, Jeanna Smialek, published [an article](https://www.nytimes.com/2020/08/27/business/economy/federal-reserve-inflation-jerome-powell.html) in the New York Times detailing the Fed's plan to deal with unemployment caused by the COVID-19 pandemic. [The article](https://www.nytimes.com/2020/08/27/business/economy/federal-reserve-inflation-jerome-powell.html) explained how the Federal Reserve is taking an unusual approach to prioritize strong labor markets over lower inflation rates. 

This visualization reproduces the historical data on interest rates.

## Clean Up The Data

```{r}
fed_rate_cleaned <- 
  fed_rate %>% 
  janitor::clean_names() %>% 
  mutate(
    observation_date = as.Date(observation_date),
    month = month(observation_date),
    year = year(observation_date),
    date = day(observation_date),
  )
```

## Plot the data

```{r}
label_point <- 
  fed_rate_cleaned %>%  
  filter(observation_date == ymd('2020-07-01')) %>% 
  mutate(fedfunds_per = scales::percent(fedfunds, scale = 1, accuracy = 0.01))
```


```{r}
recession_years <- c(
  as.Date('1990-08-01', origin = lubridate::origin),
  as.Date('2001-08-01', origin = lubridate::origin),
  as.Date('2008-01-01', origin = lubridate::origin),
  as.Date('2020-02-01',origin = lubridate::origin)
)
```


```{r}
fed_rate_cleaned %>% 
  ggplot(aes(observation_date, fedfunds)) +
  geom_vline(xintercept = recession_years, color = 'grey95', size = 5) +
  geom_line(inherit.aes = FALSE, data = fed_rate_cleaned, aes(observation_date, fedfunds), color = "#eb7154", size = 0.7) +
  geom_point(data = label_point, color = "#eb7154", size = 1.3) +
  geom_text(data = label_point, aes(label = fedfunds_per, fontface = 'bold'), hjust = -0.1, size = 3, vjust = -0.5) +
  scale_y_continuous(
    breaks = scales::breaks_width(2),
    minor_breaks = NULL,
    labels = scales::label_percent(scale = 1, accuracy = 1, suffix = "")
  ) +
  coord_cartesian(xlim = c(as.Date('1986-01-01'), as.Date('2021-01-01')), ylim = c(0.42, 9.6)) + 
  scale_x_date(
    breaks = c(as.Date('1985-01-01'), as.Date('1990-01-01'), as.Date('1995-01-01'), as.Date('2000-01-01'), as.Date('2005-01-01'), as.Date('2010-01-01'), as.Date('2015-01-01'), as.Date('2020-02-10')), 
    date_labels = "%Y",
    minor_breaks = NULL
  ) +
  annotate(
    "text",
    x = as.Date("1990-05-01"),
    y = 1,
    label = "RECESSIONS",
    size = 2,
    color = "grey40"
  ) +
  
  theme_minimal() +
  theme(
    axis.ticks.x = element_line(color = 'black'),
    axis.ticks.length.x = unit(0.20, 'cm'),
    axis.line.x = element_line(color = 'black'),
    axis.line.x.top = element_line(color = "black"),
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = 'bold')
  ) +
  labs(
    y = NULL,
    x = NULL,
    title = 'How Interest Rates Have Moved',
    subtitle = 'Interest rates are near zero as the Federal Reserve tries to support the economy while the\npandemic-induced downturn drags on.',
    caption = "Source: Federal Reserve\nTsion Tesfaye\nMotivated by The New York Times"
  )
```

